Gradual Soundness: Lessons from Static Python

### 2022-01-11 pj-7.1/paper.tex
pj: add missing $
pj: fb -> 'Meta'
pj: split sec 2 into types & runtime

%% TODO before submission
%% - look out for PROGRESSIVE TYPES~\cite{pqk-onward-2012}, where changing annotations
%%   helps the static checker find new errors. ChkDict is one example.
%% - choose words: static vs typed, nonstatic vs dynamic vs untyped
%% - use ChkDict not ChkList ... avoid confusion
%%   ... or at least mention both, be clear the focus is Dict b/c older
%% - do the LESSONS from SP come across clearly? beware making this a "progress report"
%% - terms: sound generic / concrete ... aren't people going to forget concrete? better keep as last resort

@@

\newcommand{\spapp}[2]{#1\,(#2)}
\newcommand{\spann}[2]{#1\!:\!#2}
\newcommand{\bmname}[1]{\textbf{#1}}
\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\codefont}[1]{\emph{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{T}}
\newcommand{\spteval}{\typefont{S}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptX}{\typefont{X}} %% undefined type for an early example
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptbool}{\typefont{Bool}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptdyn}{\typefont{Dynamic}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{None}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptunion}[1]{\paramtype{\typefont{Union}}{#1}}
\newcommand{\sptrawpydict}{\typefont{PyDict}}
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}
\newcommand{\sptenv}{\typefont{Env}}
\newcommand{\sptenvnil}{\cdot}
\newcommand{\sptenvempty}{\sptenvnil}
\newcommand{\sptvardef}[2]{\spann{#1}{#2}}
\newcommand{\sptfundef}[3]{\spapp{#1}{#2}\code{ -> }#3}
\newcommand{\sptclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\trule}[1]{\textsc{#1}}

@@

%\usepackage{cleveref}
\usepackage{amsthm}
\usepackage{mathpartir}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\usepackage[backend=biber]{biblatex}
\addbibresource{bg.bib}

@@

\author[a]{Kuang-Chen Lu}
\authorinfo{(\email{LuKuangchen1024@gmail.com}) is a PhD student at Brown University.}
\affiliation[a]{Brown University, Providence, RI, USA}
\author[a]{Ben Greenman}
\authorinfo{(\email{benjamin.l.greenman@gmail.com}) is a PLT member and a postdoc at Brown University.}
\author[b]{Carl Meyer}
\authorinfo{(\email{carljm@fb.com}) FILL}
\affiliation[b]{Facebook, Inc.}
\author[b]{Dino Viehland}
\authorinfo{(\email{dinoviehland@fb.com}) FILL}
\author[a]{Shriram Krishnamurthi}
\authorinfo{(\email{shriram@brown.edu}) is the Vice President of Programming Languages (no, not really) at Brown University.}

@@

\paperdetails{
  %% perspective options are: art, sciencetheoretical, scienceempirical, engineering.
  perspective=scienceempirical,
  %% State one or more areas, separated by a comma. (see 2.2)
  %% Please see list of areas in http://programming-journal.org/cfp/
  %% The list is open-ended, so use other areas if yours is/are not listed.
  area={Database programming, General-purpose programming, Program
    verification, Programming education},
  %% License options include: cc-by (default), cc-by-nc
  % license=cc-by,
}

@@

\section{Typing for Performance}
\label{s:intro}

@@

\section{\SP{} and the Cinder JIT}
\label{s:tour}
%% purpose = SP looks like Pyre, adds soundness to certain types

@@

\SP{} is a type checker and bytecode compiler.
It validates programs written in ordinary Python syntax
and generates type-specific instructions for the Cinder runtime.
The type system can express common Python idioms and is guaranteed
sound through run-time checks, which in turn enable optimizations.
The type system also includes a dynamic type as an escape hatch to
untyped behavior.
An unannotated variable behaves exactly as it would in Python.

Cinder is an extension of CPython 3.8 that adds a method-based
JIT compiler, virtual method tables, and several bytecode instructions
to directly express type checks and type-based optimizations.
These enhancements take full advantage of \SP{} type information.
They are built on top of CPython so that untyped code runs on
the platform.

Using \SP{} and Cinder is as easy as installing a new version
of CPython.
The main executable compiles programs to bytecode on the fly,
same as CPython, allowing standard developer tools to work.
By contrast, the only effective way to get type-directed optimizations
without changing CPython is to use C extension modules,
which introduce an extra compilation step and get in the way
of tooling~(\cref{s:related}).

\begin{figure}
  %% TODO replace with a real program
%  \begin{minipage}[t]{0.5\columnwidth}
  \begin{verbatim}
    from __static__ import PyDict

    def f(x: PyDict[str, int]):
      return x["A"]

    f({"A": 1}) # ==> 1
  \end{verbatim}
%  \end{minipage}\begin{minipage}[t]{0.5\columnwidth}
%  \begin{verbatim}
%    from __static__ import PyDict
%
%    def f(x: PyDict[str, int]):
%      check_args(x, PyDict)
%      return cast(x["A"], int)
%
%    invoke_function(f, {"A": 1}) # ==> 1
%  \end{verbatim}
%  %  ....
%  %  CHECK_ARGS
%  %  ....
%  %  INVOKE_FUNCTION
%  \end{minipage}
%  \begin{verbatim}
%    from __static__ import CheckedDict
%
%    def f(x: CheckedDict[str, int]):
%      return x["A"]
%
%    f(CheckedDict({"A": 1}))
%  \end{verbatim}
  \caption{A first \SP{} program}
  \label{fig:sp-example}
\end{figure}

\Cref{fig:sp-example} presents a first example program.
It defines a function \code{f} and calls it from typed code.
The syntax is normal Python with type annotations.
Only the import statement makes special use of \SP{};
it imports a type \code{PyDict} that describes Python dictionaries.

Although this program is typed, a different untyped module is free
to import the function \code{f} and invoke it with any sort of
argument.
For this reason, the compiled version of \cref{fig:sp-example} must
include a few run-time casts:
\begin{itemize}
  \item
    The function \code{f} must check that its arguments match the \code{PyDict} domain annotation.
    \SP{} inserts a check that accepts any Python dictionary.
  \item
    Because the domain check does not validate the elements of an incoming dictionary,
    the dict access (\code{x["A"]}) must check that its result matches the function codomain.
\end{itemize}
These checks ensure a \emph{shape-level} notion of type soundness~(\cref{s:model}).
In short: if an expression has type \code{T} and reduces to a value, then the value
matches the top-level constructor, or shape, of the type.
The shape for a Python dictionary is merely \code{dict}.
Other types have deeper shapes; for example, \SP{} includes a \code{ChkDict}
type that predicts the keys and values in a special kind of dictionary.

Thanks to these checks, \SP{} can leverage type soundness to generate efficient code.
Because the call to \code{f} in \cref{fig:sp-example} appears in typed code,
there is no need to check that it sends type-correct inputs.
Thus, the compiled call is optimized to use a Cinder bytecode instruction
that skips the domain check and immediately enters the function body.

The checks and optimizations in this example illustrate the general balancing act
of \SP{}.
Type soundness requires run-time checks, but also enables optimizations.
The gamble is that the optizations help more that the checks hurt.

\subsection{Type System Highlights}

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Nom~\cite{mt-oopsla-2017,mt-oopsla-2021} and PEP 484~[CITE] are two notable sources.
Relative to these, the \SP{} takes some interesting turns
because of the engineering context at Instagram.
The following contextual points are important to bear in mind as we present
notable aspects of the type system:
\begin{itemize}
  \item
    Performance is the bottom line.
    Unless \SP{} helps production code run faster, the team will be disbanded
    and reallocated.
    %% 2021-12-14: too extreme? and what about soundness?
  \item
    %% Q. why is the python so nominal? style guide? java background?
    The codebase depends heavily on first-order functions and objects in
    performance-critical code.
    Higher-order code is fairly common, but rarely appears on a critical path.
  \item
    Prior to investing in \SP{}, Instagram used Cython to compile
    performance-critical code to C extension modules.
    Cython performed well, but mixing pure Python with a few compiled modules
    made for an awkward developer experience.
    \SP{} aims to match the Python workflow.
  \item
    The Instagram codebase is already typed.
    Developers use Pyre [CITE], an optional type checker for Python,
    to write type annotations and to drive IDE tools (e.g., autocompletion).
\end{itemize}

\subsubsection{Type Dynamic}
\label{s:type-dynamic}

@@

In addition to serving as a default, the dynamic type also replaces any undefined types.
For example, suppose that the type $\sptX{}$ is undefined in \SP{} and that a
program contains the annotation $\sptchkdict{\sptint}{\sptX{}}$, which describes a
dictionary with integer keys and $\sptX{}$-typed values.
\SP{} interprets this annotation as \sptchkdict{\sptint}{\sptdyn}$ and allows any sort
of values inside the dictionary at runtime.

@@

\begin{figure}
  \begin{subfigure}[t]{0.5\columnwidth}
  \begin{verbatim}
    class A:
      def m1(self)->int:
        return 0

    class B(A):
      def m1(self):
        return 0
    # Error: dynamic cannot override int
  \end{verbatim}
    \caption{SGG Violation: removing a type in the subclass \code{B} raises a static error}
    \label{f:gg-failure-stat}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.5\columnwidth}
  \begin{verbatim}
    from __static__ import ChkDict

    def f(x: ChkDict[str, dyn]):
      return x["A"]

    d = ChkDict[str, int]({"A": 1})
    f(d)
    # Error: f expected ChkDict[str, dyn]
  \end{verbatim}
    \caption{DGG Violation: removing part of the type for the parameter \code{x} raises a dynamic error}
    \label{f:gg-failure-dyn}
  \end{subfigure}
  \caption{\SP{} maintains neither the static (SGG) nor the dynamic (DGG) gradual guarantees}
  \label{fig:gg-failure}
\end{figure}

@@

\subsubsection{Concrete Types and Shallow Types}
\label{s:checked-type}

@@

\subsubsection{Gradual Class Hierarchies}
\label{s:inheritance}

@@

\subsubsection{Tailored Types}
\label{s:idiomatic-types}

\SP{} is selective about which types to support and which to omit.
In general, supported types are widely-used and come with an efficient
implementation.
Other widely-used types are deferred to Pyre for static checks until
the day when \SP{} adds support.\footnote{At present, the \SP{} team
has no plans to support all Pyre types. The top priority is to implement
core types efficiently.}

The core supported types describe basic data (integers, booleans, strings),
data structures (lists, dicts, promises),
and nominal classes.
Additional ``C types'' enable aggressive compilation of primitive
values~(\cref{s:c-types}).
Union types are well-supported statically and narrow down in accordance
with type tests (a form of occurrence typing~\cite{tf-icfp-2010,gsk-esop-2011}),
but are hardly supported at run-time.
Only binary unions of a type and the none value get enforced with a run-time check
($\sptunion{\sptype, \sptnone}$, or $\sptoptional{\sptype}$).

Three unsupported types bear special mention.
There are no first-class class or object types, as noted in \cref{s:inheritance}.
There are also no types for first-class functions (callables).
Pyre can check these types instead, to point out clear logical flaws.
Finally, \SP{} does not support recursive types due to low demand.
Neither Pyre nor mypy have recursive types either.
%% demand seems low ... baffling
%% mypy = https://github.com/python/mypy/issues/731
%% nothing in pep 484 = https://www.python.org/dev/peps/pep-0484/

\subsection{Runtime System Highlights}
%% Non-Highlights:
%% - shadow frames = JIT feature, lite array to track Python call stack

Cinder, the \SP{} runtime system, extends the standard Python runtime in
several ways.
Some of these ensure the integrity of types.
Others use type soundness to generate efficient code.

%% TODO need this para?
We present these in less detail than the type system highlights because
we suspect they are less-widely applicable than the type system features.
Some rely on an optimizing JIT compiler, for example.
Nevertheless, these highlights are an important part of what makes \SP{} successful.

\subsubsection{Bytecode Instructions}

\subsubsection{Virtual Method Tables}
\label{s:vtable}

\subsubsection{Checked Data Structures}
\label{s:chkdict-impl}

\subsubsection{Unboxed Primitive}
\label{s:c-types}

%% 2021-12-29: TODO
%% Using the term “ctypes” for these might be confusing, since the Python stdlib
%% has a “ctypes” module that isn’t related to our primitive types. We do call the
%% base type for them CType in our compiler, but when discussing them we usually
%% call them “primitive types” or “machine types” — “C types” could also work but
%% I wouldn’t combine it into one word :)
%%
%% I just did a very rough grep check and found that we have only 12 modules
%% currently using primitive types: all int32 and int64 and cbool. So they are
%% used pretty sparingly, but they are also definitely critical to our ability to
%% match Cython in a few of the  hottest code paths that used to be Cython.

%% TODO move prim.types up?

@@

\subsubsection{Strict Modules}
\label{s:strict-mod}

\subsubsection{Classloader}
\label{s:classloader}

@@

@@

\subsection{The Payoff: Issues Reported}

While modelling \SP{}, we found several issues with the language
and submitted them as bug reports to the \SP{} developers.
Overall, we reported 25 issues~(listed in \cref{a:github-issues}).
Four of these were critical soundness issues, one of which resulted in
a segmentation fault.
All but one of these soundness issue have been fixed.
Five issues were relatively minor; these dealt with confusing error messages
and incorrect tests.
The remaining 16 issues report bugs in language design and implementation.

@@

@@

Optimizing type systems for Python ...
Reticulated;
mypyc;
Developer experience is one of the main advantages of \SP{} over mypyc,
which implements type-directed optimizations for CPython
by compiling source code to C extension modules.

@@

\begin{itemize}
  \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/LuKC1024} (N=20)
  \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/bennn} (N=4)
  \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/64} (N=1)
\end{itemize}

>>>>

%% TODO before submission
%% - look out for PROGRESSIVE TYPES~\cite{pqk-onward-2012}, where changing annotations
%%   helps the static checker find new errors. ChkDict is one example.
%% - choose words: static vs typed, nonstatic vs dynamic vs untyped
%% - use ChkDict not ChkList ... avoid confusion
%%   ... or at least mention both, be clear the focus is Dict b/c older
%% - do the LESSONS from SP come across clearly? beware making this a "progress report"
%%   - emphasize the PAYOFF from modeling early on, don't bury in sec 3
%% - terms: sound generic / concrete ... aren't people going to forget concrete? better keep as last resort
%% - perhaps a compatibility story, programmers can work on Python 3.9 until Cinder catches up b/c source code is the same

@@

\newcommand{\spapp}[2]{#1\,(#2)}
\newcommand{\spann}[2]{#1\!:\!#2}
\newcommand{\bmname}[1]{\textbf{#1}}
\newcommand{\typefont}[1]{\mathsf{#1}}
\newcommand{\codefont}[1]{\emph{#1}}
\newcommand{\paramtype}[2]{#1[#2]}
\newcommand{\sptype}{\typefont{T}}
\newcommand{\spteval}{\typefont{S}}
\newcommand{\sptclass}{\typefont{C}}
\newcommand{\sptX}{\typefont{X}} %% undefined type for an early example
\newcommand{\sptint}{\typefont{Int}}
\newcommand{\sptbool}{\typefont{Bool}}
\newcommand{\sptfloat}{\typefont{Float}}
\newcommand{\sptdyn}{\typefont{Dyn}}
\newcommand{\sptobject}{\typefont{Object}}
\newcommand{\sptnone}{\typefont{None}}
\newcommand{\sptinstanceof}[1]{\paramtype{\typefont{Instance}}{#1}}
\newcommand{\sptoptional}[1]{\paramtype{\typefont{Optional}}{#1}}
\newcommand{\sptunion}[1]{\paramtype{\typefont{Union}}{#1}}
\newcommand{\sptrawpydict}{\typefont{PyDict}}
\newcommand{\sptrawchkdict}{\typefont{ChkDict}} %% not a real SP type, but useful for TeX
\newcommand{\sptpydict}[2]{\paramtype{\sptrawpydict}{#1, #2}}
\newcommand{\sptchkdict}[2]{\paramtype{\typefont{ChkDict}}{#1, #2}}
\newcommand{\sptenv}{\typefont{Env}}
\newcommand{\sptenvnil}{\cdot}
\newcommand{\sptenvempty}{\sptenvnil}
\newcommand{\sptvardef}[2]{\spann{#1}{#2}}
\newcommand{\sptfundef}[3]{\spapp{#1}{#2}\code{ -> }#3}
\newcommand{\sptclassdef}[4]{\mathrm{class}~\spapp{#1}{#2}:~#3;~#4}
\newcommand{\trule}[1]{\textsc{#1}}

@@

\usepackage[framemethod=TikZ]{mdframed}
\newmdenv[roundcorner=2pt,linewidth=0.4pt,linecolor=Black]{spcodebox}
\usepackage{amsthm}
\usepackage{mathpartir}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\usepackage[backend=biber]{biblatex}
\addbibresource{bg.bib}

@@

\author[a]{Kuang-Chen Lu}
\authorinfo{(\email{LuKuangchen1024@gmail.com}) is a PhD student at Brown University.}
\affiliation[a]{Brown University, Providence, RI, USA}
\author[a]{Ben Greenman}
\authorinfo{(\email{benjamin.l.greenman@gmail.com}) is a PLT member and a postdoc at Brown University.}
\author[b]{Carl Meyer}
\authorinfo{(\email{carljm@fb.com}) FILL}
\affiliation[b]{Meta}
\author[b]{Dino Viehland}
\authorinfo{(\email{dinoviehland@fb.com}) FILL}
\author[a]{Shriram Krishnamurthi}
\authorinfo{(\email{shriram@brown.edu}) is the Vice President of Programming Languages (no, not really) at Brown University.}

@@

\paperdetails{
  %% perspective options are: art, sciencetheoretical, scienceempirical, engineering.
  perspective=scienceempirical,
  %% State one or more areas, separated by a comma. (see 2.2)
  %% Please see list of areas in http://programming-journal.org/cfp/
  %% The list is open-ended, so use other areas if yours is/are not listed.
  area={General-purpose programming},
  %% License options include: cc-by (default), cc-by-nc
  % license=cc-by,
}

@@

\section{Introduction}
%% alt title:
%%  {A Staged Gradual Language}
%%  {Typing for Performance}
\label{s:intro}

@@

\section{A Programmer's View of \SP{}}
\label{s:tour}
%% purpose = SP looks like Pyre, adds soundness to certain types

@@

\SP{} is part of a large codebase that includes a type system, a tailored
bytecode, and a method-based JIT compiler.
In essence, it is a new language.
The interface that it offers to programmers, however, is a familiar one.
\SP{} can handle all untyped\footnote{Python 3.8 code that uses type
annotations may, of course, contain type errors.} Python 3.8 programs, and it
compiles code on-the-fly to give the same workflow and enable the same
developer tools as standard Python.

Indeed, the advanced features of \SP{} are offered strictly on an opt-in basis.
Programmers whose decide to use its new APIs---in particular, for the type system
and the JIT compiler---can usuaally benefit without imposing maintenance costs
on their co-workers.

\begin{figure}
  \begin{subfigure}[t]{0.22\columnwidth}\begin{spcodebox}\noindent\begin{verbatim}
def f(x):
  return x["A"]

f({"A": 1})
  \end{verbatim}
  \end{spcodebox}
  %\caption{TODO}
  %\label{f:TODO}
  \end{subfigure}~\begin{subfigure}[t]{0.36\columnwidth}\begin{spcodebox}\noindent\begin{verbatim}
from __static__ import PyDict
def f(x: PyDict[Str, Int]):
  return x["A"]

f({"A": 1})
  \end{verbatim}
  \end{spcodebox}
  %\caption{TODO}
  %\label{f:TODO}
  \end{subfigure}~\begin{subfigure}[t]{0.36\columnwidth}\begin{spcodebox}\noindent\begin{verbatim}
from __static__ import ChkDict
def f(x: ChkDict[Str, Int]):
  return x["A"]

f(chkdict[Str, Int]({"A": 1}))
  \end{verbatim}
  \end{spcodebox}
  %\caption{TODO}
  %\label{f:TODO}
  \end{subfigure}

%%
%%%  \begin{minipage}[t]{0.5\columnwidth}
%%  \begin{verbatim}
%%    from __static__ import PyDict
%%    def f(x: PyDict[str, int]):
%%      return x["A"]
%%
%%    f({"A": 1}) # ==> 1
%%  \end{verbatim}
%  \end{minipage}\begin{minipage}[t]{0.5\columnwidth}
%  \begin{verbatim}
%    from __static__ import PyDict
%
%    def f(x: PyDict[str, int]):
%      check_args(x, PyDict)
%      return cast(x["A"], int)
%
%    invoke_function(f, {"A": 1}) # ==> 1
%  \end{verbatim}
%  %  ....
%  %  CHECK_ARGS
%  %  ....
%  %  INVOKE_FUNCTION
%  \end{minipage}
%  \begin{verbatim}
%    from __static__ import CheckedDict
%
%    def f(x: CheckedDict[str, int]):
%      return x["A"]
%
%    f(CheckedDict({"A": 1}))
%  \end{verbatim}
  \caption{A first \SP{} program and two migrations}
  \label{fig:sp-example}
\end{figure}

\Cref{fig:sp-example} presents a first example program and two modified versions that
utilize \SP{} types.
The basic program defines a function \code{f} and calls it with a dictionary
value.
\SP{} can run this program as-is, and even JIT-compile the function.
The other two versions import dictionary types from the \SP{} standard library:
\begin{itemize}
  \item
    The $\sptrawpydict$ type describes normal Python dictionaries.
    At compile-time, \SP{} uses this type to check the program for logical errors.
    At run-time, \SP{} enforces the type by checking that all inputs to \code{f} are
    dictionary values.
    These checks enable optimizations within the function body.

  \item
    The $\sptrawchkdict$ type describes a dictionary data structure that is new to \SP{}.
    Unlike Python dictionaries, these checked dictionaries are guaranteed to
    contain well-typed keys and values even if they escape to untyped code.
    As the body of \code{f} illustrates, the syntax for using a checked dict is standard.
    Creating a checked dict, however, requires a parameterized constructor call.
\end{itemize}
These typed versions demonstrate the multi-level nature of the \SP{} interface.
At one level, there are types that describe standard Python values that can be added
to a program with little-to-no code changes.
Beyond that, \SP{} offers special-purpose types with stronger guarantees that impose
a higher maintenance cost.

\subsection{Type System Design Goals}

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Nom~\cite{mt-oopsla-2017,mt-oopsla-2021} and PEP 484~[CITE] are two notable sources.
Relative to these, the \SP{} takes some interesting turns
because of the engineering context at Instagram.
The following contextual points are important to bear in mind:
\begin{itemize}
  \item
    Performance is the bottom line.
    Unless \SP{} helps production code run faster, the team will be disbanded
    and reallocated.
    %% 2021-12-14: too extreme? and what about soundness?
  \item
    %% Q. why is the python so nominal? style guide? java background?
    The codebase depends heavily on first-order functions and objects in
    performance-critical code.
    Higher-order code is fairly common, but rarely appears on a critical path.
  \item
    Prior to investing in \SP{}, Instagram used Cython to compile
    performance-critical code to C extension modules.
    Cython performed well, but mixing pure Python with a few compiled modules
    made for an awkward developer experience.
    \SP{} aims to match the Python workflow.
  \item
    The Instagram codebase is already typed.
    Developers use Pyre [CITE], an optional type checker for Python,
    to write type annotations and to drive IDE tools (e.g., autocompletion).
\end{itemize}
%
In general, the types that \SP{} implements all enable significant optimizations
and are simple to validate with run-time checks.
Types that do meet these criteria can be deferred to Pyre, even if they are widely-used.\footnote{
At present, the \SP{} team
has no plans to support all Pyre types. The top priority is to implement
core types efficiently.}

The core supported types describe basic data (integers, booleans, strings),
data structures (lists, dicts, promises),
and nominal classes.
Union types are well-supported statically and narrow down in accordance
with type tests (a form of occurrence typing~\cite{tf-icfp-2010,gsk-esop-2011}),
but are, for the most part, not supported at run-time.
Only binary unions of a type and the none value get enforced with a run-time check
($\sptunion{\sptype, \sptnone}$, or $\sptoptional{\sptype}$).

Three other unsupported types bear special mention.
There are no first-class class or object types~(\cref{s:inheritance}).
There are also no types for first-class functions (callables).
Pyre can check these types instead.
Third, \SP{} does not support recursive types.
Neither Pyre nor mypy have recursive types either.
%% demand seems low ... baffling
%% mypy = https://github.com/python/mypy/issues/731
%% nothing in pep 484 = https://www.python.org/dev/peps/pep-0484/

\subsection{Type Dynamic}
\label{s:type-dynamic}

@@

In addition to serving as a default, the dynamic type also replaces any undefined types.
For example, suppose that the type $\sptX{}$ is undefined in \SP{} and that a
program contains the annotation $\sptchkdict{\sptint}{\sptX{}}$, which describes a
dictionary with integer keys and $\sptX{}$-typed values.
\SP{} interprets this annotation as $\sptchkdict{\sptint}{\sptdyn}$ and allows any sort
of values inside the dictionary at runtime.

@@

\begin{figure}
  \begin{subfigure}[t]{0.5\columnwidth}
  \begin{spcodebox}
  \begin{verbatim}
    class A:
      def m1(self) -> Int:
        return 0

    class B(A):
      def m1(self):
        return 0
    # Error: Dyn cannot override Int
  \end{verbatim}
  \end{spcodebox}
    \caption{SGG Violation: removing a type in the subclass \code{B} raises a static error}
    \label{f:gg-failure-stat}
  \end{subfigure}
  ~
  \begin{subfigure}[t]{0.5\columnwidth}
  \begin{spcodebox}
  \begin{verbatim}
    from __static__ import ChkDict

    def f(x: ChkDict[Str, Dyn]):
      return x["A"]

    d = ChkDict[Str, Tnt]({"A": 1})
    f(d)
    # Error: f expected ChkDict[Str, Dyn]
  \end{verbatim}
  \end{spcodebox}
    \caption{DGG Violation: removing part of the type for the parameter \code{x} raises a dynamic error}
    \label{f:gg-failure-dyn}
  \end{subfigure}
  \caption{\SP{} maintains neither the static (SGG) nor the dynamic (DGG) gradual guarantees}
  \label{fig:gg-failure}
\end{figure}

@@

\subsection{Concrete Types and Shallow Types}
\label{s:checked-type}

@@

\subsection{Gradual Class Hierarchies}
\label{s:inheritance}

@@

\subsection{Strict Modules}
\label{s:strict-mod}

\subsection{Unboxed Primitive}
\label{s:c-types}

%% 2021-12-29: TODO
%% Using the term “ctypes” for these might be confusing, since the Python stdlib
%% has a “ctypes” module that isn’t related to our primitive types. We do call the
%% base type for them CType in our compiler, but when discussing them we usually
%% call them “primitive types” or “machine types” — “C types” could also work but
%% I wouldn’t combine it into one word :)

@@

\section{Runtime System Highlights}
%% alt title: Runtime System Highlights
%% Non-Highlights:
%% - shadow frames = JIT feature, lite array to track Python call stack

The \SP{} runtime system, formally known as Cinder, extends
CPython 3.8 in several ways to take full advantage of static types.
Cinder includes tailored bytecode instructions, virtual method tables, new
datatypes, a registry of typed modules, and a JIT compiler.

The key ingredients in Cinder may be of interest to other teams seeking to add
sound gradual typing to a widely-used language.
For example, a team of Node developers may wish to fork that runtime system
and experiment with bytecode instructions that rely on statci types.

\subsection{Bytecode Instructions}

\subsection{Virtual Method Tables}
\label{s:vtable}

\subsection{Checked Data Structures}
\label{s:chkdict-impl}

\subsection{Classloader}
\label{s:classloader}

@@

\subsection{Method-at-a-Time JIT}

FILL Cinder includes a JIT compiler for its bytecode.
The typical way to configure is to supply a list of functions to be JIT compiled.
Compilation works on any \SP{} function whether or not it is typed
(thus there is some incentive to use \SP{} even without adding type annotations).
For the biggest payoff, though, the JIT needs type information.

@@

\subsection{The Payoff: Issues Reported}
%% 2022-01-11 FILL TODO reword, excitement!

While modelling \SP{}, we found several issues with the language
and submitted them as bug reports to the \SP{} developers.
Overall, we reported 25 issues~(listed in \cref{a:github-issues}).
Four of these were critical soundness issues, one of which resulted in
a segmentation fault.
All but one of these soundness issue have been fixed.
Five issues were relatively minor; these dealt with confusing error messages
and incorrect tests.
The remaining 16 issues report bugs in language design and implementation.
The \SP{} team has acknowledged these bugs as such via GitHub a label (\code{sp-correctness}).

@@

%% 2022-01-11 TODO how many Checked modules in production?
%% 2022-01-11 TODO how many strict modules in production?
%% 2022-01-11 TODO talk about primitive types:
%% Carl: "I just did a very rough grep check and found that we have only 12 modules
%% currently using primitive types: all int32 and int64 and cbool. So they are
%% used pretty sparingly, but they are also definitely critical to our ability to
%% match Cython in a few of the  hottest code paths that used to be Cython."

@@

Optimizing type systems for Python ...
Reticulated;
mypyc;
Developer experience is one of the main advantages of \SP{} over mypyc,
which implements type-directed optimizations for CPython
by compiling source code to C extension modules.
\SP{} even used Cython at some point and found the mixed-workflow codebase painful.

@@

\begin{itemize}
  \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/LuKC1024} (N=20)
  \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/created\_by/bennn} (N=4)
  \item \shorturl{https://}{github.com/facebookincubator/cinder/issues/64} (N=1)
  %% extra, can't remember if we inspired Dino or not: https://github.com/facebookincubator/cinder/issues/32
\end{itemize}
